Navigating Through Crisis: The Influence of COVID-19 on Canada’s Mental Health Problem

### 2024-04-25 paper/paper.qmd
Fine tuning

The COVID-19 pandemic did not only threaten people's physical health but also potentially harm their mental health. In Canada, the impact of the pandemic on mental health has been profound. The Centre for Addiction and Mental Health [@CAMH] has identified a significant increase in mental health problems among Canadians, noting that the pandemic has led to unprecedented levels of anxiety and depression . Additionally, @PHAC reports on the elevated symptoms of anxiety and depression during the pandemic, further underscoring the need for comprehensive data analysis on this issue .

@@

The datasets contain some variables in common, including "REF_DATE" for time, "GEO" for location,  "Sex" or "Gender", alongside "Indicators" and "VALUE" which detail specific mental health metrics and their quantitative assessments. Apart from these, "Age" can be found in the Mental Health Indicators Dataset and The Perceived Mental Health by Age Group Dataset. The Perceived Mental Health by Gender and Sociodemographic Characteristics Dataset has a special variable "Sociodemographic_characteristics", which includes a range of detailed factors such as age, identities, race etc.

@@

The results from the analyzed datasets reveal significant trends in mental health problems across Canada. Initially, the data show a gentle increase in reported mental health concerns from the early 2000s up until 2010. Post-2010, there is a noticeable dip, followed by a steady climb in mental health issues leading up to 2015. However, the most striking feature of this trend is the sharp increase in mental health concerns starting from around 2020, which corresponds with the onset of the COVID-19 pandemic.

@@

The data also discussed the intersectional nature of mental health challenges, revealing how gender, sexuality, age, and ethnicity intersect to influence mental well-being. For instance, the increasing rates of mental health concerns among LGBTQ2+ populations stated the reality of social stressors. This leads to a discussion on systemic issues like discrimination and access to care.

@@

Despite these insights, the study has its constraints. The varied data collection methodologies over time introduce complexity in direct comparisons. Additionally, self-reported data on mental health may be influenced by evolving societal attitudes toward mental health, potentially impacting trend analysis.

@@

For future research, to comprehensively assess the lasting impact of the COVID-19 pandemic on mental health, researchers could use longitudinal studies. These studies would track individuals over extended periods, providing insights into the long-term effects of the pandemic on various demographic groups.

>>>>

The COVID-19 pandemic has not only threatened people's physical health but also potentially harmed their mental health. In Canada, the impact of the pandemic on mental health has been profound. The Centre for Addiction and Mental Health [@CAMH] has identified a significant increase in mental health problems among Canadians, noting that the pandemic has led to unprecedented levels of anxiety and depression . Additionally, @PHAC reports on the elevated symptoms of anxiety and depression during the pandemic, further underscoring the need for comprehensive data analysis on this issue .

@@

The datasets share several common variables, including 'REF_DATE' for time, 'GEO' for location, and 'Sex' or 'Gender', along with 'Indicators' and 'VALUE', which detail specific mental health metrics and their quantitative assessments. Apart from these, "Age" can be found in the Mental Health Indicators Dataset and The Perceived Mental Health by Age Group Dataset. The Perceived Mental Health by Gender and Sociodemographic Characteristics Dataset has a special variable "Sociodemographic_characteristics", which includes a range of detailed factors such as age, identities, race etc.

@@

The analysis of these datasets reveals significant trends in mental health issues across Canada. Initially, the data show a gentle increase in reported mental health concerns from the early 2000s up until 2010. Post-2010, there is a noticeable dip, followed by a steady climb in mental health issues leading up to 2015. However, the most striking feature of this trend is the sharp increase in mental health concerns starting from around 2020, which corresponds with the onset of the COVID-19 pandemic.

@@

The data also discuss the intersectional nature of mental health challenges, revealing how gender, sexuality, age, and ethnicity intersect to influence mental well-being. For instance, the increasing rates of mental health concerns among LGBTQ2+ populations stated the reality of social stressors. This leads to a discussion on systemic issues like discrimination and access to care.

@@

Despite these insights, the study faces certain constraints. Varied data collection methodologies over time introduce complexities in making direct comparisons. Moreover, self-reported data on mental health may be influenced by evolving societal attitudes, which could potentially impact trend analyses. Furthermore, the quantitative focus might not fully capture the complexities of individuals' mental health experiences, underscoring the need for qualitative approaches to deepen our understanding. These limitations indicate that future research should incorporate methodological enhancements to ensure a more comprehensive analysis.

@@

For future research, researchers could use longitudinal studies to comprehensively assess the lasting impact of the COVID-19 pandemic on mental health. These studies would track individuals over extended periods, providing insights into the long-term effects of the pandemic on various demographic groups.
